---
title: "MLB Hitting Production"
author: "Patrick Ward"
date: "12/24/2022"
output: 
  html_document:
    toc: TRUE
    number_sections: FALSE
    toc_float: TRUE
    toc_collapsed: FALSE
    theme: lumen
---

# Report Details {.tabset}

**This report provides a view into the hitting performance of MLB players drafted 2010 and on who have played in at least 8 seasons.**


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, message=FALSE)
suppressWarnings({
  suppressPackageStartupMessages({
    library(crosstalk)
    library(plotly)
    library(tidyverse)
    library(Lahman)
  })
})

### Data
df <- Batting %>%
  left_join(People) %>%
  left_join(Fielding %>% select(playerID, yearID, POS)) %>%
  mutate(first_season = substr(debut, start = 1, stop = 4),
         name = paste(nameLast, nameFirst, sep = ", ")) %>%
  filter(first_season >= 2010) %>%
  mutate(obp = (H + BB + HBP) / (AB + BB + HBP + SF),
         batting_avg = H / AB) %>%
  select(playerID, name, yearID, POS, AB, batting_avg, obp) %>%
  filter(AB >= 100) %>%
  group_by(playerID) %>%
  mutate(experience = row_number()) %>%
  arrange(playerID) %>%
  group_by(playerID) %>%
  filter(n() >= 8)
```


## Player Production

### Rookie Season Production

* Select the rookie season of interest
* Place your cursor over a dot to see which player it is

```{r}
#### Rookie season performance plot
## set highlighting capabilities for rookie season stats
rookie_key <- highlight_key(df %>%
                              filter(experience == 1),
                            ~yearID)

## create a widget for selecting rookie seasons
rookie_widget <- bscols(
  widths = 5,
  filter_checkbox("yearID", "Season", rookie_key, ~yearID, inline = TRUE)
)

rookie_plt <- bscols(
  widths = c(10),
  rookie_widget,
  plot_ly(rookie_key, x = ~batting_avg, y = ~obp, showlegend = FALSE) %>%
    add_markers(text = ~name,
                marker = list(color = '#56B4E9',
                              size = 15)) %>% 
    layout(yaxis = list(title = "OBP")) %>%
    layout(xaxis = list(title = "Batting Avg")) %>%
    layout(title = "Rookie Season Production")   
)

rookie_plt
```


### Player Time Series

* Select a player of interest to observe their trend relative to the population
* Add additional players to compare their performance and change color so that the individuals stand out


```{r}
#### Player time series plot
# player highlight key
player_key <- highlight_key(df, ~name)

# player time series plot using the highlight key
player_ts <- plot_ly(player_key, 
                     color = I("black"),
                      line = list(width = 5)) %>%
  group_by(name) %>%
  add_lines(x = ~yearID,
            y = ~batting_avg) %>% 
  layout(yaxis = list(title = "Batting Average")) %>%
  layout(xaxis = list(title = "Season"), autosize = FALSE) %>%
  layout(title = "Career Batting Average Trend")   

# player plot with interactive capability
player_plt <- highlight(
  player_ts,
  selectize = TRUE,
  dynamic = TRUE,
  persistent = TRUE
)

player_plt
```


## Distribution of performance by position

* Select the position of interest to see on base percentage
* Change colors when adding new positions to allow for comparison

```{r}
#### Position histogram
# position highlight key
pos_key <- highlight_key(df, ~POS)

# BA histogram
ba_hist <- plot_ly(pos_key, 
                   color = I("light grey")) %>%
  group_by(POS) %>%
  add_histogram(x = ~batting_avg) %>%
  layout(xaxis = list(title = "Batting Avg"), autosize = FALSE)

# OBP histogram
obp_hist <- plot_ly(pos_key, 
                   color = I("light grey")) %>%
  group_by(POS) %>%
  add_histogram(x = ~obp) %>%
  layout(xaxis = list(title = "OBP"), autosize = FALSE)

subplot(ba_hist, 
        obp_hist, 
        nrows = 1, 
        shareY = TRUE,
        titleX = TRUE) %>%
  layout(barmode = "overlay",
         showlegend = FALSE) %>%
  highlight(dynamic = TRUE,
            selectize = TRUE,
            persistent = TRUE,
            selected = attrs_selected(opacity = 0.5))
```

