---
title: "Two Group Comparison - Frequentist vs Bayes"
author: "Patrick Ward"
date: "5/29/2022"
output: html_document
---

## Two Groups

We are conducting a study on the effect of a new strength training program. Participants are randomly allocated into a control group (n = 17), receiving a normal training program, or an experimental group (n = 22), receiving the new training program.

The data consists of the change in strength score observed for each participant in their respective groups. (NOTE: For the purposes of this example, the strength score is a made up number describing a participants strength).

**Simulate the data**

```{r}
library(tidyverse)
library(patchwork)

theme_set(theme_classic())

set.seed(6677)
df <- data.frame(
  subject = 1:39,
  group = c(rep("control", times = 17), rep("experimental", times = 22)),
  strength_score_change = c(round(rnorm(n = 17, mean = 0, sd = 0.5), 1),
                       round(rnorm(n = 22, mean = 0.5, sd = 0.5), 1))
) %>%
  mutate(group = factor(group, levels = c("experimental", "control")))

df

```


## Summary Statistics

```{r}
df %>%
  group_by(group) %>%
  summarize(N = n(),
            Avg = mean(strength_score_change),
            SD = sd(strength_score_change),
            SE = SD / sqrt(N))
```


It looks like the new strength training program led to a greater improvement in strength, on average, than the normal strength training program.


## Plot the data

**Density plots of the two distributions**

```{r}
df %>%
  ggplot(aes(x = strength_score_change, fill = group)) +
  geom_density(alpha = 0.2) +
  xlim(-2, 2.5)
```


**Plot the means and 95% CI to compare visually**

```{r}
df %>%
  group_by(group) %>%
  summarize(N = n(),
            Avg = mean(strength_score_change),
            SD = sd(strength_score_change),
            SE = SD / sqrt(N)) %>%
  ggplot(aes(x = group, y = Avg)) +
  geom_hline(yintercept = 0,
             size = 1,
             linetype = "dashed") +
  geom_point(size = 5) +
  geom_errorbar(aes(ymin = Avg - 1.96 * SE, ymax = Avg + 1.96 * SE),
                width = 0,
                size = 1.4) +
  theme(axis.text = element_text(face = "bold", size = 13),
        axis.title = element_text(face = "bold", size = 17))
```


**Plot the 95% CI for the difference in means**

```{r}
df %>%
  mutate(group = factor(group, levels = c("control", "experimental"))) %>%
  group_by(group) %>%
  summarize(N = n(),
            Avg = mean(strength_score_change),
            SD = sd(strength_score_change),
            SE = SD / sqrt(N),
            SE2 = SE^2,
            .groups = "drop") %>%
  summarize(diff = diff(Avg),
            se_diff = sqrt(sum(SE2))) %>%
  mutate(group = 'Group\nDifference') %>%
  ggplot(aes(x = diff, y = 'Group\nDifference')) +
  geom_vline(aes(xintercept = 0),
             linetype = "dashed",
             size = 1.2) +
  geom_point(size = 5) +
  geom_errorbar(aes(xmin = diff - 1.96 * se_diff,
                    xmax = diff + 1.96 * se_diff),
                width = 0,
                size = 1.3) +
  xlim(-0.8, 0.8) +
  labs(y = NULL,
       x = "Average Difference in Strengh Score",
       title = "Difference in Strength Score",
       subtitle = "Experimental - Control (Mean Difference ± 95% CI)") +
  theme(axis.text = element_text(face = "bold", size = 13),
        axis.title = element_text(face = "bold", size = 17),
        plot.title = element_text(size = 20),
        plot.subtitle = element_text(size = 18))
```


**Initial Observations**

* We can see that the two distributions have a decent amount of overlap
* The mean change in strength for the experimental group appears to be larger than the mean change in strength for the control group
* When visually comparing the group means and their associated confidence intervals, there is an observable difference. Additionally, looking at the mean difference plot, that difference appears to be significant based on the fact that the confidence interval does not cross zero.

So, it seems like something is going on here with the new strength training program. Let's do some testing to see just how large the difference is between our two groups.


## Group Comparison: T-test

First, let's use a frequentist approach to evaluating the difference in strength score change between the two groups. We will use a t-test to compare the mean differences.

```{r}
diff_t_test <- t.test(strength_score_change ~ group,
       data = df,
       alternative = "two.sided")

diff_t_test

## store some of the main values in their own elements to use later
t_test_diff  <- as.vector(round(diff_t_test$estimate[1] - diff_t_test$estimate[2], 2))

se_diff <- 0.155    # calculated above when creating the plot
```


* The test is statistically significant, as the p-value is less than the magic 0.05 (p = 0.01) with a mean difference in strength score of 0.41 and a 95% Confidence Interval of 0.1 to 0.73.

Here, the null hypothesis was that the difference in strength improvement between the two groups is 0 (no difference) while the alternative hypothesis is that the difference is not 0. In this case, we did a two sided test so we are stating that we want to know if the change in strength from the new training program is either better or worse than the normal training program. If we were interested in testing the difference in a specific direction, we could have done a one sided test in the direction of our hypothesis, for example, testing whether the new training program is better than the normal program.

But, what if we just got lucky with our sample in the experimental group, and they happened to respond really well to the new program, and got unlucky in our control group, and they happened to not change as much to the normal training program? This could certainly happen. Also, had we collected data on a few more participants, it is possible that the group means could have changed and no effect was observed.

The t-test only allows us to compare the group means. It doesn't tell us anything about the differences in variances between groups (we could do an F-test for that). The other issue is that rarely is the difference between two groups exactly 0. It makes more sense for us to compare a full distribution of the data and make an inference about how much difference there is between the two groups. Finally, we are only dealing with the data we have on hand (the data collected for our study). What if we want to incorporate prior information/knowledge into the analysis? What if we collect data on a few more participants and want to add that to what we have already observed to see how it changes the distribution of our data? For that, we can use a Bayesian approach.

## Group Comparison: Bayes

The two approaches we will use are:

1) Group comparison with a known variance, using conjugate priors
2) Group comparisons without a know variance, estimating the mean and SD distributions jointly

#### Part 1: Group Comparison with known variance

First, we need to set some priors. Let's say that we are skeptical about the effect of the new strength training program. It is a new training stimulus that the subjects have never been exposed to, so perhaps we believe that it will improve strength but not much more than the traditional program. We will set our prior belief about the mean improvement of any strength training program to be 0.1 with a standard deviation of 0.3, for our population. Thus, we will look to combine this average/expected improvement (prior knowledge) with the observations in improvement that we see in our respective groups. Finally, we convert the standard deviation of the mean to `precision`, calculated as `1/sd^2`, for our equations going forward.

```{r}
prior_mu <- 0.1
prior_sd <- 0.3
prior_var <- prior_sd^2
prior_precision <- 1 / prior_var
```

Plot the prior distribution for the mean difference

```{r}
hist(rnorm(n = 1e6, mean = prior_mu, sd = prior_sd),
     main = "Prior Distribution for difference in strength change\nbetween control and experimental groups",
     xlab = NULL)
abline(v = 0,
       col = "red",
       lwd = 5,
       lty = 2)
```


Next, to make this conjugate approach work we need to have a known standard deviation. In this example, we will not be estimating the joint posterior distributions (mean and SD). Rather, we are saying that we are interested in knowing the mean and the variability around it but we are going to have a fixed/known SD.

Let's say we looked at some previously published scientific literature and also tried out the new program in a small pilot study of athletes and we the improvements of a strength training program are normally distributed with a known SD of 0.6.

```{r}
known_sd <- 0.6
known_var <- known_sd^2
```


Finally, let's store the summary statistics for each group in their own elements. We can type these directly in from our summary table.

```{r}
df %>%
  group_by(group) %>%
  summarize(N = n(),
            Avg = mean(strength_score_change),
            SD = sd(strength_score_change),
            SE = SD / sqrt(N))

experimental_N <- 22
experimental_mu <- 0.423
experimental_sd <- 0.494
experimental_var <- experimental_sd^2
experimental_precision <- 1 / experimental_var

control_N <- 17
control_mu <- 0.0118
control_sd <- 0.469
control_var <- control_sd^2
control_precision <- 1 / control_var

```


Now we are ready to update the observed study data for each group with our prior information and obtain posterior estimates. We will use the updating rules provided by William Bolstad in Chapter 13 of his book, [Introduction to Bayesian Statistics, 2nd Ed](https://www.amazon.com/Introduction-Bayesian-Statistics-William-Bolstad/dp/0470141158).

```{r}
##### Update the control group observations ######
## SD
posterior_precision_control <- prior_precision + control_N / known_var
posterior_var_control <- 1 / posterior_precision_control
posterior_sd_control <- sqrt(posterior_var_control)

## mean
posterior_mu_control <- prior_precision / posterior_precision_control * prior_mu + (control_N / known_var) / posterior_precision_control * control_mu

posterior_mu_control
posterior_sd_control

##### Update the experimental group observations ######
## SD
posterior_precision_experimental <- prior_precision + experimental_N / known_var
posterior_var_experimental <- 1 / posterior_precision_experimental
posterior_sd_experimental <- sqrt(posterior_var_experimental)

## mean
posterior_mu_experimental <- prior_precision / posterior_precision_experimental * prior_mu + (experimental_N / known_var) / posterior_precision_experimental * experimental_mu

posterior_mu_experimental
posterior_sd_experimental
```


Compare the posterior difference in strength change

```{r}
# mean
mu_diff <- posterior_mu_experimental - posterior_mu_control
mu_diff

# sd = sqrt(var1 + var2)
sd_diff <- sqrt(posterior_var_experimental + posterior_var_control)
sd_diff

# 95% Credible Interval
mu_diff + qnorm(p = c(0.025, 0.975)) * sd_diff
```


* Combining the observations for each group with our prior, it appears that the new program was more effective than the normal program on average (0.34 difference in change score between experimental and control) but the credible interval suggests the data is consistent with the new program having no extra benefit [95% Credible Interval: -0.0003 to 0.69].

Next, we can take the posterior parameters and perform a Monte Carlo Simulation to compare the posterior distributions.

Monte Carlo Simulation is an approach that uses random sampling from a defined distribution to solve a problem. Here, we will use Monte Carlo Simulation to sample from the normal distributions of the control and experimental groups as well as a simulation for the difference between the two. To do this, we will create a random draw of 10,000 values with the mean and standard deviation being defined as the posterior mean and standard deviation from the respective groups.

```{r}
## Number of simulations
N <- 10000

## Monte Carlo Simulation
set.seed(9191)
control_posterior <- rnorm(n = N, mean = posterior_mu_control, sd = posterior_sd_control)
experimental_posterior <- rnorm(n = N, mean = posterior_mu_experimental, sd = posterior_sd_experimental)
diff_posterior <- rnorm(n = N, mean = mu_diff, sd = sd_diff)

## Put the control and experimental groups into a data frame
posterior_df <- data.frame(
  group = c(rep("control", times = N), rep("experimental", times = N)),
  posterior_sim = c(control_posterior, experimental_posterior)
)

```


Plot the mean and 95% CI for both simulated groups

```{r}
posterior_df %>%
  group_by(group) %>%
  summarize(Avg = mean(posterior_sim),
            SE = sd(posterior_sim)) %>%
  ggplot(aes(x = group, y = Avg)) +
  geom_hline(yintercept = 0,
             size = 1,
             linetype = "dashed") +
  geom_point(size = 5) +
  geom_errorbar(aes(ymin = Avg - 1.96 * SE, ymax = Avg + 1.96 * SE),
                width = 0,
                size = 1.4) +
  labs(x = NULL,
       y = "Mean Diff",
       title = "Difference in Strength Score",
       subtitle = "Monte Carlo Simulation of Posterior Mean and SD") +
  theme(axis.text = element_text(face = "bold", size = 13),
        axis.title = element_text(face = "bold", size = 17),
        plot.title = element_text(size = 20),
        plot.subtitle = element_text(size = 17))
```


* We see more overlap here than we did when we were just looking at the observed data. Part of this is due to the priors that we set.


Plot the posterior simulation for the difference

```{r}
hist(diff_posterior,
     main = "Posterior Simulation of the Difference between groups",
     xlab = "Strength Score Difference")
abline(v = 0,
       col = "red",
       lwd = 5,
       lty = 2)
```


**Compare these results to the observed data and the t-test results**

```{r}
data.frame(
  group = c("control", "experimental", "difference"),
  observed_avg = c(control_mu, experimental_mu, t_test_diff),
  posterior_sim_avg = c(posterior_mu_control, posterior_mu_experimental, mu_diff),
  observed_standard_error = c(control_sd / sqrt(control_N), experimental_sd / sqrt(experimental_N), se_diff),
  posterior_sim_standard_error = c(posterior_sd_control, posterior_sd_experimental, sd_diff)
  )
```


* Recall that we set out prior mean to 0.1 and our prior SD to 0.3
* Also notice that, because this conjugate approach is directed at the means, I covert the observed standard deviations into standard errors (the standard deviation of the mean given the sample size). Remember, we set the SD to be known for this approach to work, so we are not able to say anything about it following our Bayesian procedure. To do so, we'd need to estimate both parameters (mean and SD) jointly. We will do this in part 2.

* Above, we see that, following the Bayesian approach, the mean value of the control group gets pulled up closer to our prior while the mean value of the experimental group gets pulled down closer to that prior.
* The posterior for the difference between the two groups, which is what we are interested in, has come down towards the prior, considerably. Had we had a larger sample of data in our two groups, the movement towards the prior would have been less extreme. But, since this was a relatively small study we get more movement towards the prior, placing less confidence in the observed sample of participants we had. We also see that the standard error around the mean difference increased, since our prior SD was 0.3. So, the mean difference decreased a bit and the standard error around that mean increased a bit. Both occurrences lead to less certainty in our observed outcomes.

**Compare the Observed Difference to the Bayesian Posterior Difference**

* Start by creating a Monte Carlo Simulation of the observed difference from the original data.
* Notice that the Bayesian Posterior Difference (blue) is pulled down towards our prior, slightly.
* We also see that the Bayesian Posterior has more overlap of 0 than the Observed Difference (red) from the original data, which had a small sample size and hadn't combined any prior knowledge that we might have had.

```{r}
N <- 10000

set.seed(8945)
observed_diff <- rnorm(n = N, mean = t_test_diff, sd = se_diff)

plot(density(observed_diff),
     lwd = 5,
     col = "red",
     main = "Comparing the Observed Difference (red)\nto the\nBayes Posterior Difference (blue)",
     xlab = "Difference in Strength Score between Groups")
lines(density(diff_posterior),
      lwd = 5,
      col = "blue")
abline(v = 0,
       lty = 2,
       lwd = 3,
       col = "black")
```


#### Part 2: Group Comparison - Joint estimation of mean and SD

We are comparing the differences in a change in strength score (pre to post) between groups: a group that had a normal training program and a group that was given a new/experimental strength training program.

Review the data set.

```{r}
df %>%
  head()
```


In Part 1, we compared the difference in outcomes between groups using a t-test (frequentist) and a Bayesian approach, where we took advantage of a normal conjugate distribution. In order to use the normal-normal conjugate, we needed to make an assumption about a known population standard deviation. By using a known standard deviation it meant that we only needed to perform Bayesian updating for the mean of the distribution, allowing us to compare between group means and their corresponding standard errors. The problem with this approach is that we might not always have a known standard deviation to apply, thus we would want to be able to estimate this along with the mean -- we need to estimate both parameters, jointly!

To do this in the Bayesian framework we will work with a regression model where `group` (control or experimental) is the independent variable and the dependent variable is the change in strength score. We can do this because, recall, [t-tests are really just regression underneath](http://optimumsportsperformance.com/blog/t-test-anova-its-linear-regression-all-the-way-down/).

Let's look at the output of a simple linear regression before trying a Bayesian regression.

```{r}
fit_lm <- lm(strength_score_change ~ group, data = df)
summary(fit_lm)
confint(fit_lm)

t.test(strength_score_change ~ group, data = df)
```


As expected, the results that we get here are the same as those that we previously obtained from our t-test. The coefficient for the control group (-0.411) represents the difference in the mean strength score compared to the experimental group, whose mean change in strength score is accounted for in the intercept.

Because the experimental group is now represented as the model intercept, we can instead code the model without an intercept and get a mean change in strength score for both groups. This is accomplished by adding a `0` to the right side of the equation, telling `R` that we don't want a model intercept.

```{r}
fit_lm2 <- lm(strength_score_change ~ 0 + group, data = df)
summary(fit_lm2)
confint(fit_lm2)
```


Okay, now let's move into the Bayesian framework. We'll utilize the help of the brilliant `brms` package, which compiles `C++` and runs the `Stan` language under the hood, allowing you to use the simple and friendly `R` syntax that you're used to.

Let's start with a simple Bayesian Regression Model.

```{r}
library(brms)

# Set 3 cores for parallel processing (helps with speed)
fit_bayes1 <- brm(strength_score_change ~ group, 
                 data = df,
                 cores = 3,
                 seed = 7849
                 )

summary(fit_bayes1)
```


The output here is a little more extensive than what we are used to with the normal regression output. Let's make some notes:

* The control group coefficient still represents the mean difference in strength score compared to the experimental group.
* The experimental groups mean strength score is still the intercept
* The coefficients for the intercept and control group are the same that we obtained with the normal regression.
* We have a new parameter at the bottom, `sigma`, which is a value of 0.50. This value represents the shared standard deviation between the two groups. If you recall the output of our frequentist regression model, we had a value called `residual standard error`, which was 0.48 (pretty similar). The one thing to add with our `sigma` value here is, like the model coefficients, it has its own error estimate and 95% Credible Intervals (which we do not get from the original regression output).

Before going into posterior simulation, we have to note that we only got one `sigma` parameter. This is basically saying that the two groups in our model are sharing a standard deviation. This is similar to running a t-test with equal variances (**NOTE: the default in R's `t-test()` function is `var.equal = FALSE`, which is usually a safe assumption to make.**). To specify a `sigma` value for both groups we will wrap the equation in the `bf()` function, which is a function for specifying `brms` formulas. In there, we will indicate different sigma values for each group to be estimated. Additionally, to get a coefficient for both groups (versus the experimental group being the intercept), we will add a `0` to the right side of the equation, similar to what we did in our second frequentist regression model above.

```{r}
group_equation <- bf(strength_score_change ~ 0 + group,
                     sigma ~ 0 + group)

fit_bayes2 <- brm(group_equation, 
                 data = df,
                 cores = 3,
                 seed = 7849
                 )

summary(fit_bayes2)
```


Now we have an estimate for each group (their mean change in strength score from pre to post testing) and a sigma value for each group(**NOTE: To get this value to the normal scale we need to take is exponential as they are on a log scale, as indicated by the `links` statement at the top of the model output, `sigma = log`.**). Additionally, we have credible intervals around the coefficients and sigmas.

```{r}
exp(-0.69)
exp(-0.72)
```


We have not specified any priors yet, so we are just using the default priors. Before we try and specify any priors, let's get posterior samples from our model (don't forget to exponentiate the sigma values). We will also calculate a Cohen's d as a measure of standardized effect.

$Cohens.d = (group.diff) / sqrt((group1.sd^2 + group2.sd^2) / 2)$

```{r}
bayes2_draws <- as_draws_df(fit_bayes2) %>%
  mutate(across(.cols = contains("sigma"), 
                ~exp(.x)),
         group_diff = b_groupexperimental - b_groupcontrol,
         cohens_d = group_diff / sqrt((b_sigma_groupexperimental^2 + b_sigma_groupcontrol^2)/2))

bayes2_draws %>%
  head()
```


Let's make a plot of the difference in means and Cohen's d across our 4000 posterior samples.

```{r}
par(mfrow = c(1,2))
hist(bayes2_draws$group_diff,
     main = "Posterior Draw of Group Differences",
     xlab = "Group Differences")
abline(v = 0,
       col = "red",
       lwd = 3,
       lty = 2)
hist(bayes2_draws$cohens_d,
     main = "Posterior Draw of Cohen's d",
     xlab = "Cohen's d")
```


Okay, now let's add some priors and repeat the process of plotting the posterior samples. We will use the same normal prior for the means that we used in Part 1, Normal(0.1, 0.3) and for the sigma value we will use a Cauchy prior, Cauchy(0, 1).

```{r}
## fit model
fit_bayes3 <- brm(group_equation, 
                 data = df,
                 prior = c(
                       set_prior("normal(0.1, 0.3)", class = "b"),
                       set_prior("cauchy(0, 1)", class = "b", dpar = "sigma")
                 ),
                 cores = 3,
                 seed = 7849
                 )

summary(fit_bayes3)

## exponent of the sigma values
exp(-0.66)
exp(-0.70)

## posterior draws
bayes3_draws <- as_draws_df(fit_bayes3) %>%
  mutate(across(.cols = contains("sigma"), 
                ~exp(.x)),
         group_diff = b_groupexperimental - b_groupcontrol,
         cohens_d = group_diff / sqrt((b_sigma_groupexperimental^2 + b_sigma_groupcontrol^2)/2))

bayes3_draws %>%
  head()

## plot sample of group differences and Cohen's d
par(mfrow = c(1,2))
hist(bayes3_draws$group_diff,
     main = "Posterior Draw of Group Differences",
     xlab = "Group Differences")
abline(v = 0,
       col = "red",
       lwd = 3,
       lty = 2)
hist(bayes3_draws$cohens_d,
     main = "Posterior Draw of Cohen's d",
     xlab = "Cohen's d")
```


Combine all of the results together so we can evaluate what has happened.

```{r}
no_prior_sim_control_mu <- mean(bayes2_draws$b_groupcontrol)
no_prior_sim_experimental_mu <- mean(bayes2_draws$b_groupexperimental)
no_prior_sim_diff_mu <- mean(bayes2_draws$group_diff)

no_prior_sim_control_sd <- sd(bayes2_draws$b_groupcontrol)
no_prior_sim_experimental_sd <- sd(bayes2_draws$b_groupexperimental)
no_prior_sim_diff_sd <- sd(bayes2_draws$group_diff)

with_prior_sim_control_mu <- mean(bayes3_draws$b_groupcontrol)
with_prior_sim_experimental_mu <- mean(bayes3_draws$b_groupexperimental)
with_prior_sim_diff_mu <- mean(bayes3_draws$group_diff)

with_prior_sim_control_sd <- sd(bayes3_draws$b_groupcontrol)
with_prior_sim_experimental_sd <- sd(bayes3_draws$b_groupexperimental)
with_prior_sim_diff_sd <- sd(bayes3_draws$group_diff)


data.frame(
  group = c("control", "experimental", "difference"),
  observed_avg = c(control_mu, experimental_mu, t_test_diff),
  posterior_sim_avg = c(posterior_mu_control, posterior_mu_experimental, mu_diff),
  no_prior_sim_avg = c(no_prior_sim_control_mu, no_prior_sim_experimental_mu, no_prior_sim_diff_mu),
  with_prior_sim_avg = c(with_prior_sim_control_mu, with_prior_sim_experimental_mu, with_prior_sim_diff_mu),
  observed_standard_error = c(control_sd / sqrt(control_N), experimental_sd / sqrt(experimental_N), se_diff),
  posterior_sim_standard_error = c(posterior_sd_control, posterior_sd_experimental, sd_diff),
  no_prior_sim_standard_error = c(no_prior_sim_control_sd, no_prior_sim_experimental_sd, no_prior_sim_diff_sd),
  with_prior_sim_standard_error = c(with_prior_sim_control_sd, with_prior_sim_experimental_sd, with_prior_sim_diff_sd)
  ) %>%
  mutate(across(.cols = -group,
                ~round(.x, 3))) %>%
  t() %>%
  as.data.frame() %>%
  setNames(., c("Control", "Experimental", "Difference")) %>%
  slice(-1) %>%
  mutate(models = rownames(.),
    group = c("Average", "Average", "Average", "Average", "Standard Error", "Standard Error", "Standard Error", "Standard Error")) %>%
  relocate(models, .before = Control) %>%
  group_by(group) %>%
  gt::gt()
```

Let's make some notes:

* First, observed refers to the actual observed data, posterior_sim is our normal-normal conjugate (using a known standard deviation), no_prior_sim is our Bayesian regression with default priors and with_prior_sim is our Bayesian regression with pre-specified priors.

* In the normal-normal conjugate (posterior_sim) analysis (Part 1), both the control and experimental groups saw their mean values get pulled closer to the prior leading to a smaller between group difference than we saw in the observed data.

* The Bayesian regression with no priors specified (no_prior_sim) resulted in a mean difference that is pretty much identical to the outcome we saw with our t-test on the observed data.

* The Bayesian Regression with specified priors (with_prior_sim) ends up being somewhere in the middle of the observe data/Bayes Regression with no priors and the normal-normal conjugate. The means for both groups are pulled close to the prior but not as much as the normal-normal conjugate means (posterior_sim). Therefore, the mean difference between groups is higher than the posterior_sim output but not as large as the observed data (because it is influenced by our prior). Additionally, the group standard errors are more similar to the observed data with the Bayesian regression with priors than the Bayesian regression without priors and the normal-normal Bayesian analysis.
