---
title: "Skill vs Luck"
author: "Patrick Ward"
date: "12/3/2022"
output: html_document
---

## Load libraries and data

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(gt)

## Get data from "pro_sport_win_tables.xlsx"
premier_lg <- readxl::read_excel("pro_sport_win_tables.xlsx", sheet = 'premier league')
nhl <- readxl::read_excel("pro_sport_win_tables.xlsx", sheet = 'nhl')
nba <- readxl::read_excel("pro_sport_win_tables.xlsx", sheet = 'nba')
nfl <- readxl::read_excel("pro_sport_win_tables.xlsx", sheet = 'nfl')
mlb <- readxl::read_excel("pro_sport_win_tables.xlsx", sheet = 'mlb')

```


**From The Success Equation by Michael Mauboussin (pg. 78)**

$Variance(Observed) = Variance(Skill) + Variance(Luck)$


Since we can estimate the Variance(Observed), which is the standard deviation of win% across a league, and we can estimate the Variance(Luck), which is the standard deviation of the league win%, with a little bit of algebra, the above equation becomes:

$Variance(Skill) = Variance(Observed) - Variance(Luck)$

**However, what we care about here is the contribution of luck to winning. This is calculated via the following ratio:**

$Contribution.of.Luck = Variance(Luck) / Variance(Observed)$


Create a function for these calculations

```{r}
league_perf <- function(avg_win_pct, obs_sd, luck_sd, league){
  
  ## convert standard deviations to variance
  var_obs <- obs_sd^2
  var_luck <- luck_sd^2
  
  ## calculate the variation due to skill
  var_skill <- var_obs + var_luck
  
  ## calculate the contribution of luck to winning
  luck_contribution <- var_luck / var_obs
  
  
  ## Results table
  output <- tibble(
    league = {{league}},
    avg_win_pct = avg_win_pct,
    luck_contribution = luck_contribution,
  )
  
  return(output)
  
}
```



## Premier League Soccer

```{r}
## get info for function
# NOTE: Treating Draws as 1/3 of a win, since that reflects the points the team is awarded in the standings
premier_lg %>%
  select(Squad, MP, W, D) %>%
  mutate(win_pct = (W + 1/3*D) / MP) %>%
  summarize(games = max(MP),
            avg_win_pct = mean(win_pct),
            obs_sd = sd(win_pct),
            luck_sd = sqrt((avg_win_pct * (1 - avg_win_pct)) / games))

## Run the function
premier_lg_output <- league_perf(avg_win_pct = 0.462,
                                 obs_sd = 0.155,
                                 luck_sd = 0.0809,
                                 league = "Premier League")

premier_lg_output
```

## NHL

```{r}
## get info for function
nhl %>%
  select(Season, Tm, GP, W) %>% 
  mutate(win_pct = W / GP) %>%
  summarize(games = max(GP),
            avg_win_pct = mean(win_pct),
            obs_sd = sd(win_pct),
            luck_sd = sqrt((avg_win_pct * (1 - avg_win_pct)) / games))

## Run the function
nhl_output <- league_perf(avg_win_pct = 0.5,
                                 obs_sd = 0.116,
                                 luck_sd = 0.0552,
                                 league = "NHL")

nhl_output

## 2008 - 2012 NHL Results
nhl2008_2012 <- readxl::read_excel("2008-2012 NHL Results.xlsx")

nhl2008_2012 %>%
  select(team, GP, W) %>% 
  mutate(win_pct = W / GP) %>%
  summarize(games = max(GP),
            avg_win_pct = mean(win_pct),
            obs_sd = sd(win_pct),
            luck_sd = sqrt((avg_win_pct * (1 - avg_win_pct)) / games))

nhl2008_2012_output <- league_perf(avg_win_pct = 0.5,
           obs_sd = 0.0799,
           luck_sd = 0.0552,
           league = "NHL 2008-2012")

nhl2008_2012_output
```


## NBA

```{r}
## get info for function
nba %>%
  select(season, team, W, L, win_pct) %>% 
  summarize(games = max(W + L),
            avg_win_pct = mean(win_pct),
            obs_sd = sd(win_pct),
            luck_sd = sqrt((avg_win_pct * (1 - avg_win_pct)) / games))

## Run the function
nba_output <- league_perf(avg_win_pct = 0.498,
                                 obs_sd = 0.143,
                                 luck_sd = 0.0552,
                                 league = "NBA")

nba_output
```


## NFL

```{r}
## get info for function
nfl %>%
  select(season, Tm, W, L, win_pct) %>% 
  summarize(games = max(W + L),
            avg_win_pct = mean(win_pct),
            obs_sd = sd(win_pct),
            luck_sd = sqrt((avg_win_pct * (1 - avg_win_pct)) / games))

## Run the function
nfl_output <- league_perf(avg_win_pct = 0.5,
                                 obs_sd = 0.194,
                                 luck_sd = 0.121,
                                 league = "NFL")

nfl_output
```


## MLB

```{r}
## get info for function
mlb %>%
  filter(season != 2020) %>%
  select(season, Tm, W, L, win_pct) %>% 
  summarize(games = max(W + L),
            avg_win_pct = mean(win_pct),
            obs_sd = sd(win_pct),
            luck_sd = sqrt((avg_win_pct * (1 - avg_win_pct)) / games))

## Run the function
mlb_output <- league_perf(avg_win_pct = 0.5,
                                 obs_sd = 0.0916,
                                 luck_sd = 0.0393,
                                 league = "MLB")

mlb_output

## 2007 - 2011 MLB Results
mlb2007_2011 <- readxl::read_excel("2007-2011 MLB Results.xlsx")

mlb2007_2011 %>%
  select(Tm, W, L, win_pct) %>% 
  summarize(games = 162,
            avg_win_pct = mean(win_pct),
            obs_sd = sd(win_pct),
            luck_sd = sqrt((avg_win_pct * (1 - avg_win_pct)) / games))

## Run the function
mlb2007_2011_output <- league_perf(avg_win_pct = 0.5,
                                 obs_sd = 0.0661,
                                 luck_sd = 0.0392,
                                 league = "MLB 2007-2011")

mlb2007_2011_output

```


## Results Table

```{r}
bind_rows(
  premier_lg_output,
  nhl_output,
  nba_output,
  nfl_output,
  mlb_output
) %>%
  arrange(luck_contribution) %>%
  gt(rowname_col = "league") %>%
  cols_label(
    avg_win_pct = md("**Lg Avg Win%**"),
    luck_contribution = md("**% Contribution of\nLuck on Winning**")
  ) %>%
  fmt_percent(columns = avg_win_pct:luck_contribution,
              decimals = 1)

## historic data
bind_rows(
  nhl2008_2012_output,
  mlb2007_2011_output
) %>%
  arrange(luck_contribution) %>%
  gt(rowname_col = "league") %>%
  cols_label(
    avg_win_pct = md("**Lg Avg Win%**"),
    luck_contribution = md("**% Contribution of\nLuck on Winning**")
  ) %>%
  fmt_percent(columns = avg_win_pct:luck_contribution,
              decimals = 1)
```

